---
title: "vaccine"
author: "May Wu PID:A59010588"
date: "11/24/2021"
output:
  pdf_document:
   latex_engine: xelatex
---

## Mini-Project

# COVID-19 Vaccination Rates

```{r}
vax <- read.csv( 'covid19vaccinesbyzipcode_test.csv' )
head(vax)
```

> Q. how many entries do we have?

```{r}
nrow(vax)
```

we can use the **skimr** package and the `skim()` function to get a quick overview of the data

```{r}
skimr::skim(vax)
```

Notice that one of these columns is a date column. Working with time and dates gets annoying quickly. we can use **lubridate** package to make this easy `install.packages("lubridate")`

```{r}
library(lubridate)
today()

```

> how many days since the first entry in the dataset?

```{r}
vax$as_of_date[1]
```

This will not work because our data column was read as characters
```{r}
#today()-vax$as_of_date[1]
```

```{r}
d = ymd(vax$as_of_date)
```

```{r}
today()-d[1]
```
will make the `as_of_date` column Date format

```{r}
vax$as_of_date = ymd(vax$as_of_date)
```

> Q. When was the dataset last updated? what is the last date in this dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```
> Q. how many days since the last update?

```{r}
 vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q. how many different zip code

```{r}
length(unique(vax$zip_code_tabulation_area))
```
To work with zipcode, we can use the **zipcodeR**
```{r}
library(zipcodeR)
reverse_zipcode(c('92037', "92109") )
```
## Focus on San Diego county

subset the full CA `vax` data down to just SD County using `filter()`

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

More complicated subsetting 

```{r}

sd.20 = filter(vax, county == "San Diego",
       age5_plus_population > 20000)
nrow(sd.20)
```

> Q. What is the average vaccination rate of SD as of yesterday?

```{r}
sd.now = filter(vax, county == "San Diego", as_of_date == '2021-11-23')
head(sd.now)
```

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

> Q. make a histogram of these value

base R histogram
```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

> Q. What's the average vaccination value for 92037 area?

```{r}
d = filter(sd.now,zip_code_tabulation_area == '92037' )
d$percent_of_population_fully_vaccinated
```

> Q. what's the population of the 92037 ZIP code area?

```{r}
d$persons_fully_vaccinated
```

```{r}
d = filter(sd.now,zip_code_tabulation_area == '92124' )
d$percent_of_population_fully_vaccinated
```

time series of vaccination rate for a given ZIP code area.

starting with 92037


```{r}

lj = filter(vax, zip_code_tabulation_area == '92037')
library(ggplot2)
#vax$percent_of_population_fully_vaccinated
ggplot(lj, aes(x=as_of_date, y=percent_of_population_fully_vaccinated)) + geom_line(group = 1) + geom_point() + labs(x='Date', y = 'percent vaccinated')

```
Let's make this plot for the whole SD county

```{r}
sd.36 = filter(vax, county == 'San Diego', age5_plus_population > 36144)
head(sd.36)
```

> Q. how many ZIP code areas in SD county have a population larger than 92037

```{r}
length(unique(sd.36$zip_code_tabulation_area))
```
lets make the plot

```{r}
ggplot(sd.36, aes(x=as_of_date, y=percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) ) + 
  geom_line() + 
  labs(x='Date', y = 'percent vaccinated')

```

> Q. Make a plot like this for the all ZIP code areas in the State with a population at least as large as La Jolla

```{r}
ca = filter(vax,  age5_plus_population > 36144)
head(ca)
```

```{r}
length(unique(ca$zip_code_tabulation_area))
```
```{r}
ggplot(ca, aes(x=as_of_date, y=percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) ) + 
  geom_line(alpha = 0.2) + 
  labs(x='Date', y = 'percent vaccinated') + 
  geom_hline(yintercept = 0.67, color = 'red')

```

> Q. What is the mean across the state for these 36k+ population area?

```{r}
ca.now = filter(ca, as_of_date =='2021-11-23')
summary(ca.now$percent_of_population_fully_vaccinated)

```




